---
title: "PHW251: Midterm Live Session"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#run libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(lubridate)
library(kableExtra)
```

We will be using data from the Tidy Tuesday project. Tidy Tuesday is a weekly social data project available for free online. They have a goal of "making learning to work with data easier" and are a great resource for interesting data sets. To learn more about this project and explore the many data sets they have available, go to <https://github.com/rfordatascience/tidytuesday>.

For these practice problems, we will be using the "bob_ross" data set, which is a data set of all of the paintings painted on Bob Ross's TV show, "The Joy of Painting." A full data dictionary is available at the source website (<https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-02-21>). For our purposes, we are interested in the following variables:

| Variable       | Description                                                          |
|----------------|----------------------------------------------------------------------|
| painting index | Painting number as enumerated in collection.                         |
| painting title | Title of the painting.                                               |
| Season         | Season of 'The Joy of Painting' in which the painting was featured.  |
| Episode        | Episode of 'The Joy of Painting' in which the painting was featured. |
| num_colors     | Number of unique colors used in the painting.                        |
| colors         | List of colors used in the painting.                                 |
| Cadmium_Yellow | Cadmium_Yellow used                                                  |

1.  First, read in the file provided, "bob_ross.csv".

```{r}
#read in file
#data comes from tidy tuesday (https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-02-21). 

#YOUR CODE HERE

#ANSWERS
bob_ross <- read_csv("bob_ross.csv")
```

2.  Use functions you learned in Week 3 to investigate what kind of data you have here. What functions would you use to do this? Information you might want to know is: how many observations are there? how many variables? what kind of data?

```{r, echo=F}
#YOUR CODE HERE

#ANSWERS
dim(bob_ross)
str(bob_ross)
glimpse(bob_ross)
 #403 observations, 27 variables.

```

3.  The column names could be more reader-friendly; their current format is not very 'tidy'.
    a.  Rename the column names so that all names are lowercase and snake case.
    b.  Use R to create an output displaying the new names.

```{r}
#YOUR CODE HERE

#ANSWERS
bob_ross <- bob_ross %>%
  rename_with(~ tolower(gsub(" ","_",.x,fixed=TRUE))) 

names(bob_ross)
```

4.  There are a lot of variables here and we just want to focus on a few of them.

    a.  Create a new data frame called bob_ross_abbr that includes only the following variables:

    -   painting_title
    -   season
    -   episode
    -   num_colors
    -   colors

    b.  Save the data frame as a tibble.

```{r}
#YOUR CODE HERE

#ANSWERS
bob_ross_abbr <- bob_ross %>%
  select(painting_title, season, episode, num_colors, colors)

as_tibble(bob_ross_abbr)
```

5.  You want to categorize the number of colors used for each painting in the bob_ross_abbr data frame.

    a.  Use a case_when() statement to add a new variable "color_cat" that categorizes a painting based on the number of colors (num_colors) used.
    b.  Categorize as follows: "low" if 2-5 used, "medium" if 6-8 colors used, and "high" if 9 or more colors used. Any cases with just one color of paint should be categorized as "outlier".
    c.  Arrange the observations by num_colors in ascending order before taking a look at bob_ross_abbr again.

```{r}
#YOUR CODE HERE

#ANSWERS
bob_ross_abbr <- bob_ross_abbr %>%
  mutate(color_cat = case_when(num_colors >=2 & num_colors< 6 ~ "low",
                               num_colors >=6 & num_colors<9 ~ "medium", 
                               num_colors >=9 ~ "high", 
                               TRUE ~ "outlier"
    
  )) %>%
  arrange(num_colors)
head(bob_ross_abbr)
```

6.  Could you have completed the task in the question above with an if_else() statement? Why or why not? *YOUR ANSWER HERE*

#ANSWERS *Yes, you could have used nested if_else() statements to accomplish this. Using case_when() vs. using if_else() is a matter of your preference. However, using if_else() can be more cumbersome and more difficult to debug.*

7.  With bob_ross_abbr, group by the variable, season, and then summarize to determine the average number of unique colors used per season. Save as a new data frame called bob_ross_colors.

```{r}
#YOUR CODE HERE

#ANSWERS
bob_ross_colors <- bob_ross_abbr %>% 
  group_by(season) %>% 
  summarize(avg_num_colors = mean(num_colors))
head(bob_ross_colors)
```

8.  Determine from the bob_ross_colors data frame which season Bob used the most unique colors on average. Save your answer to an object called "most_colorful_season".

Challenge: How would you use R to print out the value for this object using base R code?

```{r}
#YOUR CODE HERE

#ANSWERS
most_colorful_season <- 14
#with base R
most_colorful_season <- 
  bob_ross_colors[bob_ross_colors$avg_num_colors==max(bob_ross_colors$avg_num_colors),]$season
most_colorful_season
```

9.  Visualize what you learned in problems 7 & 8. Use the bob_ross_colors data frame to make a graph (e.g., line chart, scatterplot) showing how the average number of colors used may have changed over the seasons.

```{r}
#YOUR CODE HERE

#ANSWERS
ggplot(data = bob_ross_colors, aes(x=season, y=avg_num_colors))+
  geom_line()+
  labs(title="Bob Ross Paintings", subtitle = "Average Number of Colors Used per Season", x="Season", y="Average number of colors")+
  theme_classic()
```

10. You want to look closer at the names of the paintings from Season 1 only. Create a table using the kableExtra package that just displays the painting_title and episode number. Add a caption and column names that feel appropriate.

```{r, results='asis'}
#YOUR CODE HERE

#ANSWERS
library(kableExtra)
season_1 <- bob_ross_abbr %>% 
  filter(season == 1) %>%
  select(painting_title, episode)%>%
  arrange(episode)


knitr:: kable(season_1,
              booktabs=TRUE,
              format="simple",
              caption = "Paintings from Season 1",
              col.names = c("Painting Title", "Episode #"))
```
