---
title: "PHW251: Midterm Live Session - Student"
author: "Lyndsay Miles"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
#run libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(lubridate)
library(kableExtra)
```

We will be using data from the Tidy Tuesday project. Tidy Tuesday is a weekly social data project available for free online. They have a goal of "making learning to work with data easier" and are a great resource for interesting data sets. To learn more about this project and explore the many data sets they have available, go to <https://github.com/rfordatascience/tidytuesday>.

For these practice problems, we will be using the "bob_ross" data set, which is a data set of all of the paintings painted on Bob Ross's TV show, "The Joy of Painting." A full data dictionary is available at the source website (<https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-02-21>). For our purposes, we are interested in the following variables:

| Variable       | Description                                                          |
|-----------------|----------------------------------|
| painting index | Painting number as enumerated in collection.                         |
| painting title | Title of the painting.                                               |
| Season         | Season of 'The Joy of Painting' in which the painting was featured.  |
| Episode        | Episode of 'The Joy of Painting' in which the painting was featured. |
| num_colors     | Number of unique colors used in the painting.                        |
| colors         | List of colors used in the painting.                                 |
| Cadmium_Yellow | Cadmium_Yellow used                                                  |

1.  First, read in the file provided, "bob_ross.csv".

```{r}
#read in file
#data comes from tidy tuesday (https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-02-21). 

#YOUR CODE HERE

bob_ross <- NULL
```

2.  Use functions you learned in Week 3 to investigate what kind of data you have here. What functions would you use to do this? Information you might want to know is: how many observations are there? how many variables? what kind of data?

```{r, echo=F}
#YOUR CODE HERE


```
-   Number of observations: **YOUR ANSWER HERE**
-   Number of variables: **YOUR ANSWER HERE**

3.  The column names could be more reader-friendly; their current format is not very 'tidy'.
    a.  Rename the column names so that all names are lowercase and snake case.
    b.  Use R to create an output displaying the new names.

```{r}
#YOUR CODE HERE


```

4.  There are a lot of variables here and we just want to focus on a few of them.

    a.  Create a new data frame called bob_ross_abbr that includes only the following variables:

    -   painting_title
    -   season
    -   episode
    -   num_colors
    -   colors

    b.  Save the data frame as a tibble.

```{r}
#YOUR CODE HERE


```

5.  You want to categorize the number of colors used for each painting in the bob_ross_abbr data frame.

    a.  Use a case_when() statement to add a new variable "color_cat" that categorizes a painting based on the number of colors (num_colors) used.
    b.  Categorize as follows: "low" if 2-5 used, "medium" if 6-8 colors used, and "high" if 9 or more colors used. Any cases with just one color of paint should be categorized as "outlier".
    c.  Arrange the observations by num_colors in ascending order before taking a look at bob_ross_abbr again.

```{r}
#YOUR CODE HERE

#Hint: You can use mutate, case_when(), arrange(), and head() for this problem.

```

6.  Could you have completed the task in the question above with an if_else() statement? Why or why not? 
*YOUR ANSWER HERE*

7.  With bob_ross_abbr, group by the variable, season, and then summarize to determine the average number of unique colors used per season. Save as a new data frame called bob_ross_colors.

```{r}
#YOUR CODE HERE


bob_ross_colors <- NULL
head(bob_ross_colors)
```

8.  Determine from the bob_ross_colors data frame which season Bob used the most unique colors on average. Save your answer to an object called "most_colorful_season". 

  Challenge: How would you use R to print out the value for this object using base R code?

```{r}
#YOUR CODE HERE


most_colorful_season <- NULL


```

9.  Visualize what you learned in problems 7 & 8. Use the bob_ross_colors data frame to make a graph (e.g., line chart, scatterplot) showing how the average number of colors used may have changed over the seasons.

```{r}
#YOUR CODE HERE

```

10. You want to look closer at the names of the paintings from Season 1 only. Create a table using the kableExtra package that just displays the painting_title and episode number. Add a caption and column names that feel appropriate.

```{r, results='asis'}
#YOUR CODE HERE

```
